---
title: "Homework 2"
author: "Jake Vettoretti"
date: "2025-09-26"
output: github_document
---

```{r}
library(tidyverse) #installing necessary libraries
library(readr)
library(readxl)
```

## Problem 1

Installing and cleaning pols data.

```{r}
pols_df = read_csv("data/pols-month.csv") |> #Clean pols.csv data
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) |> 
  mutate(
    month = month.name[as.numeric(month)],
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem",
      TRUE ~ NA_character_
    )
  ) |> 
  select(-prez_dem, -prez_gop, -day) |> 
  arrange(year, month)

pols_df
```

Installing and cleaning snp data.

```{r}
snp_df = read_csv("data/snp.csv") |>  # Clean snp.csv data
  janitor::clean_names() |>
  mutate(
    date = as.Date(date, format = "%m/%d/%y"),
    year = as.integer(ifelse(
      format(date, "%y") <= "68", 
      paste0("20", format(date, "%y")),
      paste0("19", format(date, "%y"))
    )),
    month = month.name[as.integer(format(date, "%m"))]
  ) |>
  select(year, month, close) |>
  arrange(year, month)

snp_df
```

Preparing datasets for merge.

```{r}
unemployment_df = read_csv("data/unemployment.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = jan:dec,
    names_to = "month_abbr",
    values_to = "unemployment_rate"
  ) |> 
  # Convert abbreviations using month.name and month.abb constants
  mutate(month = month.name[match(month_abbr, tolower(month.abb))]) |> 
  select(-month_abbr) |> 
  rename(year = year) |> 
  arrange(year, month)
```

Merging the data sets

```{r}
merged_data = pols_df |> #merging pols and snp to prepare for merge into unemployment
  left_join(snp_df, by = c("year", "month"))

final_data = merged_data |>  # Step 2: Merge unemployment into the result  
  left_join(unemployment_df, by = c("year", "month"))

final_data
```

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset 
(e.g. give the dimension, range of years, and names of key variables):

## Problem 2

Importing and cleaning the Mr. Trash Wheel Data

```{r}
# Mr. Trash Wheel
mr_trash_wheel = read_excel("data/trashwheeldatanew.xlsx", 
  sheet = "Mr. Trash Wheel", skip = 1, na = c("NA", ".", "")) %>%  # Skip the first row (metadata/header),
  janitor:: clean_names() |>  # Clean column names and remove rows that don't have dumpster data (NA in dumpster column)
  filter(!is.na(dumpster))|>
  mutate(sports_balls = as.integer(round(sports_balls))) |> 
  select(dumpster:homes_powered)   # Select only the relevant columns with dumpster-specific data
mr_trash_wheel

# Professor Trash Wheel - same process with consistent data types
professor_trash_wheel = read_excel("data/trashwheeldatanew.xlsx", 
  sheet = "Professor Trash Wheel", skip = 1, na = c("NA", ".", "")) %>%  
  janitor::clean_names() |>  
  filter(!is.na(dumpster))|>
  # Ensure consistent data types
  mutate(
    year = as.character(year)  # Convert to character to match Mr. Trash Wheel
  ) |>
  select(dumpster:homes_powered)

# Gwynnda Trash Wheel - same process with consistent data types
gwynns_trash_wheel = read_excel("data/trashwheeldatanew.xlsx", 
  sheet = "Gwynns Falls Trash Wheel", skip = 1, na = c("NA", ".", "")) %>%  
  janitor::clean_names() |>  
  filter(!is.na(dumpster))|>
  # Ensure consistent data types
  mutate(
    year = as.character(year)  # Convert to character to match Mr. Trash Wheel
  ) |>
  select(dumpster:homes_powered)

# Add identifier variable to each dataset
mr_trash_wheel = mr_trash_wheel |> 
  mutate(trash_wheel = "Mr. Trash Wheel")

professor_trash_wheel = professor_trash_wheel |> 
  mutate(trash_wheel = "Professor Trash Wheel")

gwynns_trash_wheel = gwynns_trash_wheel |> 
  mutate(trash_wheel = "Gwynns Falls Trash Wheel")

all_trash_wheels = mr_trash_wheel |> 
  full_join(professor_trash_wheel) |> 
  full_join(gwynns_trash_wheel) |> 
  select(trash_wheel, everything())  # Move identifier to first column

# View the combined tidy dataset
all_trash_wheels
view(all_trash_wheels)

#Calculations requested for trash tons total and cigarette butts for June 2022
professor_total_weight <- all_trash_wheels |> 
  filter(trash_wheel == "Professor Trash Wheel") |> 
  summarise(total_weight_tons = sum(weight_tons, na.rm = TRUE))

gwynns_june_2022_butts <- all_trash_wheels |> 
  filter(trash_wheel == "Gwynns Falls Trash Wheel",
         year == "2022",
         month == "June") |> 
  summarise(total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE))

# Display results
cat("Total weight of trash collected by Professor Trash Wheel:\n")
print(professor_total_weight)

cat("\nTotal number of cigarette butts collected by Gwynns in June 2022:\n")
print(gwynns_june_2022_butts)
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. 
For available data, what was the total weight of trash collected by Professor Trash Wheel? 
  282 tons.
What was the total number of cigarette butts collected by Gwynnda in June of 2022?
  18,120 cigarette butts.

## Problem 3
```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(readr)

# Read the dataset
zipcodes <- read_csv("data/zillowdata/ZipCodes.csv")


# Data cleaning and transformation
cleaned_zipcodes <- zipcodes %>%
  # Convert all character columns to proper case for consistency
  mutate(across(where(is.character), ~ stringr::str_to_title(.))) %>%
  # Handle missing values in Neighborhood - replace "Na" with proper NA
  mutate(Neighborhood = ifelse(Neighborhood == "Na", NA, Neighborhood)) %>%
  # Convert File Date to proper Date format
  mutate(`File Date` = as.Date(`File Date`, format = "%m/%d/%y")) %>%
  # Ensure ZipCode is character (to preserve leading zeros if any)
  mutate(ZipCode = as.character(ZipCode)) %>%
  # Create a proper County name column (currently "County" has borough names)
  rename(Borough = County) %>%
  # Create proper FIPS code by combining State and County FIPS
  mutate(FIPS_Code = paste(`State FIPS`, `County FIPS`, sep = "")) %>%
  # Reorder columns logically
  select(
    Borough, 
    ZipCode, 
    Neighborhood,
    FIPS_Code,
    `State FIPS`, 
    `County Code`, 
    `County FIPS`,
    `File Date`
  ) %>%
  # Sort by Borough and ZipCode for better organization
  arrange(Borough, ZipCode)


# Final organized dataset
final_zipcodes <- cleaned_zipcodes %>%
  # Final arrangement for meaningful order
  arrange(Borough, Neighborhood, ZipCode)
```






